I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-04-27 lipics-master/comppaths.tex
influence sign + bias

Consider a neuron $n$ in layer $k$, for which we aim to establish accurate bounds $[\alpha,\beta]$. Define $X_n$ as the set of {\em unstable} neurons $x$ in layers up to $k-1$, where each $x$ is part of a compensating path targeting neuron $n$ (except its the source node). Let $\overline{X_n}$ represents the set of all other neurons in the first $k-1$ layers, essentially serving as the complement of $X_n$. We introduce MILP${X_n}$, a variant of the MILP encoding where all nodes in $\overline{X_n}$ are linearly relaxed (if the neuron is stable, it is the exact linear mode of the ReLU), hence incorporating $|X_n|$ binary variables, each corresponding to a neuron in $X_n$. Additionally, this encoding encapsulates bounds for every neuron $n'$ in the initial $k-1$ layers, as computed inductively in previous steps. That is, we do not use explicitly in $X_n$ nodes on compensating paths with target neurons before $n$ ({\em indirectly} compensating paths), only those with target $n$ ({directly} compensating paths).
The contribution of these {\em indirectly} compensating paths are thus 
only taken into account indirectly via the bounds on previous neurons computed inductively,
and not as binary variables. Theorem \ref{th2} states that the bounds $[\alpha,\beta]$ for neuron $n$, as computed by MILP${X_n}$, are accurate, thereby offering an algorithm for DNN verification in scenarios involving compensating paths. We will provide the proof under a fairly light well-connected hypothesis (H1), which only very specific DNNs does not satisfy, namely very sparse DNNs where almost all weights are 0. (H1) removes some corner cases, simplifying the proof.

(H1) for every 2 (not necessarily totally disjoint) paths $\rho_1 \neq \rho_2$ with non-zero weight with the same source $m$ and the same target $n$ with at least 2 transitions, there exists another path $\rho_3$ from $m$ to $n$ (with non zero-weight) totally disjoint from $\rho_1$ and $\rho_2$ (except at the source and target).

@@

The sketch of proof of Theorem \ref{th2} can be found in Section \ref{sec.proofs2}.

@@

First, notice that by continuity of all the functions used in the DNNs, it suffices to show both theorem for $\gamma=\alpha$ and $\gamma=\beta$.
%Intuitively, our first proof shows that if there are no compensation, then there are no correlations between nodes.
Consider a target neuron $z$.
In case there is no compensating path, we can assign a "sign" $S(n)$ to each neuron $n$: $S(n)=0$
if all paths from $n$ to $z$ has weight 0, $+1$ if all paths have positive weights, and 
$-1$ if they have negative weights. %Using the concept of sign, 
We introduce
% input vectors 
$\vx^{*}, \vx^{\sharp}$:

\begin{definition}
We define the following two input vectors $\vx^{*}, \vx^{\sharp}$ in $\cal B$: 
	\begin{itemize}
		\item $\vx^{*}$ is the input vector defined in the following way:
		\begin {itemize}
		 \item $x^*_{a_i}=\max(a_i)$ if $S(a_i)\in \{0,1\}$, and
          \item $x^*_{a_i}=\min(a_i)$ otherwise, that is when $S(a_i)=-1$.
	\end{itemize}

@@

We can then prove that $\vx^\star,\vx^\sharp$ maximize and minimize conjointly all the nodes in the DNN. It does {\em not} mean that there is no dependencies between nodes.
There can be correlations between nodes, but they do not affect the $\min$ and $\max$ bounds that can be reached.

@@

Hence $\val_{\vx^\star}(z)=\beta$ and $\val_{\vx^\sharp}(z)=\alpha$,
for $[\alpha, \beta]$ computed by the Box abstraction. We conclude Theorem \ref{th1} by invocking continuity. 
Notice that biases have no impact on $S(n)$ (it is {\em not} the sign of )nor the

The full proof is in appendix A.

>>>>

Consider a neuron $n$ in layer $k$, for which we aim to establish accurate bounds $[\alpha,\beta]$. Define $X_n$ as the set of {\em unstable} neurons $x$ in layers up to $k-1$, where each $x$ is part of a compensating path targeting neuron $n$ (except its the source node). Let $\overline{X_n}$ represents the set of all other neurons in the first $k-1$ layers, essentially serving as the complement of $X_n$. We introduce MILP${X_n}$, a variant of the MILP encoding where all nodes in $\overline{X_n}$ are linearly relaxed (if the neuron is stable, it is the exact linear mode of the ReLU), hence incorporating $|X_n|$ binary variables, each corresponding to a neuron in $X_n$. Additionally, this encoding encapsulates bounds for every neuron $n'$ in the initial $k-1$ layers, as computed inductively in previous steps. That is, we do not use explicitly in $X_n$ nodes on compensating paths with target neurons before $n$ ({\em indirectly} compensating paths), only those with target $n$ ({directly} compensating paths).
The contribution of these {\em indirectly} compensating paths are thus 
only taken into account indirectly via the bounds on previous neurons computed inductively,
and not as binary variables. Theorem \ref{th2} states that the bounds $[\alpha,\beta]$ for neuron $n$, as computed by MILP${X_n}$, are accurate, thereby offering an algorithm for DNN verification in scenarios involving compensating paths. We provide the proof in Section \ref{sec.proofs2}, under a fairly light well-connected hypothesis (H1), which only very sparse DNNs where almost all weights are 0 does not satisfy. (H1) removes some corner cases, simplifying the proof.

(H1) for every 2 (not necessarily totally disjoint) paths $\rho_1 \neq \rho_2$ with at least 2 non-zero transitions with the same source $m$ and target $n$, there exists another path $\rho_3$ from $m$ to $n$ (with non zero-weight) totally disjoint from $\rho_1$ and $\rho_2$ (except at the source and target).

@@

@@

First, notice that by continuity of all the functions used in the DNNs, it suffices to show both theorem for $\gamma=\alpha$ and $\gamma=\beta$.
%Intuitively, our first proof shows that if there are no compensation, then there are no correlations between nodes.
Consider a target neuron $z$.
In case there is no compensating path, we can assign an "influence sign" $S(n)$ to each neuron $n$: $S(n)=0$ if all paths from $n$ to $z$ has weight 0, $S(n)=+1$ if all paths have positive weights, and $S(n)=-1$ if they have negative weights. E.g., $S(n)=-1$ means that $\val_{\vx}(z)$ will increase if $\val_{\vx}(n)$ decreases. 
Specifically, $S(n)$ is {\em not} the sign of $\val_{\vx}(n)$, and biases have no impact on  $S(n)$: bias in $n$ changes $\val_{\vx}(n)$, not $S(n)$, while biases in between $n$ and $z$ will not change the influence of $\val_{\vx}(n)$ over $\val_{\vx}(z)$ (though it will change $\val_{\vx}(z)$ itself). We introduce $\vx^{*}, \vx^{\sharp}$:
%Using the concept of sign,

\begin{definition}
Let $\vx^{*}, \vx^{\sharp}$ be the two input vectors in $\cal B$ such that: 
	\begin{itemize}
		\item $\vx^{*}$ is the input vector defined in the following way:
		\begin {itemize}
		 \item $x^*_{a_i}=\max(a_i)$ if $S(a_i)\in \{0,1\}$, and
          \item $x^*_{a_i}=\min(a_i)$ otherwise, that is when $S(a_i)=-1$.
	\end{itemize}

@@

We can then prove that $\vx^\star,\vx^\sharp$ maximize and minimize conjointly all the nodes in the DNN. It does {\em not} mean that there is no dependencies between nodes.
There can be correlations between nodes, but they do not affect the $\min$ and $\max$ values that can be reached.

@@

Hence $\val_{\vx^\star}(z)=\beta$ and $\val_{\vx^\sharp}(z)=\alpha$,
for $[\alpha, \beta]$ computed by the Box abstraction. We conclude Theorem \ref{th1} by invocking continuity. The full proof is in appendix A.
